---
title: "Chapter 4"
output: html_document
date: "Last update: 23/02/2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Create New Variables
We always create new variables through this syntax.
```r
variable_name <- expression
```
The composition above, *expression* can involve many operators and functions, 
here are some arithmetic operators that can form formulas.
```{r table-knit-1, echo = FALSE}
library(knitr)
library(kableExtra)
library(magrittr)
data <- data.frame(
  operator = c("\\+", "\\-", "\\*", "/", "^ or **", "x %% y", "x %/% y"),
  description = c("Add",
                  "Subtract",
                  "Multiply",
                  "Divide",
                  "Exponentiation",
                  "Complementation(x mod y), 5 %% 2 = 1",
                  "Integer Division, 5 %/% 2 = 2")
)
kable(data, caption = "Arithmetic operators") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"), 
                full_width = FALSE) %>% 
    column_spec(1:2, width = "5cm", latex_column_spec = "c")
```
1. Function **transform()** can do anything useful with data frames currently. Here
is an example, we use this function to add two variables to data frame mydata.
```{r, include = TRUE}
mydata <- data.frame(x1 = c(2, 2, 6, 4), x2 = c(3, 4, 2, 8))
mydata <- transform(mydata, sumx = x1 + x2, meanx = (x1 + x2) / 2)
mydata
```
### Reencoding of Variables
