---
title: "Chapter 1"
output: html_document
date: "Last update: 16/02/2025"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### R Help System
#### Opening the help documentation
```r
help.start()   # Opens the main index of the R help system in a web browser  
help("foo")    # Shows the help page for function "foo"  
?foo           # Equivalent to help("foo")  
```
```r
help.search("foo")  # Searches "foo" in local R documentation  
??foo               # Shortcut for help.search("foo")  
```
```r
example("foo")  # Runs the examples provided in the help page of "foo"  
```
### Miscellaneous R commands
```r
# Use the up (^) and down (v) arrow keys on your keyboard to browse past commands  
setwd("mydirectory")  # Set the current working directory to "mydirectory"  
history(#)  # Display the last "#" orders, by default "#" is 25  
save(objectlist, file = "myfile") # Save the specified object to a file  
load("myfile")  # Read a workplace to the current command window  
q() # Quit R  
```
1. R use backslash(\\) as an escape character.  
2. **setwd()** will not automatically create a non-exist content. If necessary, we can 
use **dir.create()** to create a new content, then use the former function to point 
the workplace to the new one.  
3. **source("filename")** will run a script in current window.  
```r
source("myscript.R")  # This clause will run "myscript.R" here
```
4. **sink("filename")** will redirect the output to "filename". By default, if 
the file exists, then replaces it. Additionally, we can use parameter **append = TRUE** 
to add the output to the last of the file. Parameter **split = TRUE** can add the 
output to screen and file at the same time. If we call this function without 
arguments, it will add output to screen only. It should be noted that the target
address of **sink()** do not include a file name suffix and should be quoted by
double quotation marks.  
5. If there are some outputs are images, then follow the instruction.
```{r table-knitr, echo = FALSE}
library(knitr)
library(kableExtra)
library(magrittr)
data <- data.frame(
  functions = c("bmp(\"filename.bmp\")", "jpeg(\"filename.jpeg\")", 
                "pdf(\"filename.pdf\")", "png(\"filename.png\")",               "postscript(\"filename.ps\")", "svg(\"filename.svg\")", 
                "win.metafile(\"filename.wmf\")"),
  outputs = c("BMP file", "JPEG file", "PDF file", "PNG file", "Postscript file", 
        "SVG file", "Windows metafile")
)
kable(data, caption = "Functions for Saving Graphic Output") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"), 
                full_width = FALSE) %>% 
    column_spec(1:2, width = "5cm", latex_column_spec = "c")
```
6. There are many R functions can be used to manage packages. To install a new 
package, use the command **install.packages()**. The package name as a parameter
should be quoted.
7. Finally, we should use function **dev.off()** to redirect the output to terminal.